\large{\bf Looking forwards and backwards: dynamics and genealogies of locally regulated populations }

### 2023-05-02 forwards-and-backwards.tex
.

The following appears as Theorem A.2 in \citet{etheridge/kurtz:2019},
specialized slightly to the case that the processes 
are c\`adl\`ag and have no fixed points of discontinuity.
For an $S_0$-valued, measurable process $Y$, $\hat{\mathcal{F}}^Y_t$
denotes the completion of the $\sigma$-algebra generated by
$Y(0)$ and $\{\int_0^r h(Y(s)) ds, r \le t, h \in B(S_0)\}$.
Also, let $D_S[0,\infty)$ denote the space of c\`adl\`ag, $S$-valued functions
with the Skorohod topology, and $M_S[0,\infty)$ the space of Borel measurable functions
from $[0,\infty)$ to $S$,
topologized by convergence in Lesbegue measure.
For other definitions see \citet{etheridge/kurtz:2019}.

@@

In this section we verify one of the conditions of the Markov Mapping Theorem,
namely, that ``integrating out levels'' in the generator of the lookdown process
we obtain the generator of the projected process.
In the notation of the theorem,
we are verifying that $C$ defined in~\eqref{eqn:C_defn} is in fact $\Pgen^N$
(if defined with $A^N$) or $\Pgen^\infty$ (if defined with $A$).
We will work with test functions of the form
\begin{equation} \label{eqn:f_defn}
    f(\lp) = \prod_{(x, u) \in \lp} g(x, u) = \exp\left(\langle \log g, \lp \rangle \right) ,
\end{equation}
where $0 \le g \le 1$ and $g(x,u) = 1$ for all $u \ge u_g$ for some $u_g < \infty$.
Furthermore, recall that $\kappa^N(\lp)(\cdot) = \lp(\cdot \times [0, N)) / N$
is the ``spatial projection operator'',
and define the transition function $\Gamma^N : \measures \to \mathcal{M}(\IR^d \times [0,N))$
so that for $\eta \in \measures$, if $\hat g(x) = \int_0^N g(x, u) du / N$, then
\begin{align*}
    F^N_g(\eta)
    &:=
    \int f(\lp) \Gamma^N(\eta, d\lp) \\
    &=
    \exp\left(
        N \left\langle
            \log \frac{1}{N} \int_0^N g(x, u) du, \eta(dx)
        \right\rangle
    \right) \\
    &=
    \exp\left(
        N \left\langle \log \hat g(x), \eta(dx) \right\rangle
    \right) ,
\end{align*}
i.e., $\Gamma^N$ assigns independent labels on $[0, N]$ to each of the points in $\eta$.
It follows from Lemma~\ref{def: MP definition of limit} that for test functions of this form
the generator of $\eta^N_t$ is
\begin{align} \label{eqn:pgen_defn_again}
    \begin{split}
    \Pgen^N F^N_g(\eta)
    =
    F^N_g(\eta)
        N \theta & \bigg\langle
        \gamma(x, \eta) \int r(z, \eta) \left( \hat g(z) - 1 \right) q_\theta(x, dz)
    \\ &\qquad \qquad {}
        +
        \mu_\theta(x, \eta) \left( \frac{1}{\hat g(x)} - 1 \right)
        ,
        \eta(dx)
    \bigg\rangle .
    \end{split}
\end{align}
(Note that $f$ here differs from the $f$ used in Lemma~\ref{def: MP definition of limit}
so as to agree with standard usage in the literature on lookdown processes.)
The generator of $\lp^N_t$ is $A^N$, defined in equation~\eqref{eqn:lookdown_generator}.

@@

We now integrate each piece against $\Gamma^N$.
First note that by the product form of $f$,
$$
    \int f(\lp) \sum_{(x,u)\in\lp} \frac{\ell(x,u)}{g(x,u)} \Gamma^N(\eta,d\lp)
    =
    F_g^N(\eta) \bigg\langle
        \frac{1}{N \hat g(x)} \int_0^N \ell(x,u) du,
        N \eta(dx)
    \bigg\rangle .
$$
Therefore,
\footnotesize
\begin{align*}
&
    \int A^N_1f(\lp) \Gamma^N(\eta,d\lp) 
\\ &\qquad =
    F_g^N(\eta) \bigg\langle \frac{c_\theta(x,\eta)}{\hat g(x)} \int_0^N \Bigg\{
        \int_u^N \Bigg(
            \frac{1}{2} g(x,v_1) \int_{\IR^d}(g(y,u)-g(x,u))q^m_{\theta}(x,dy,\eta)
        \Bigg) dv_1
    \Bigg\} du, \eta(dx) \bigg \rangle
\\ &\qquad =
    F_g^N(\eta) \bigg\langle \frac{c_\theta(x,\eta)}{\hat g(x)} \Bigg\{
    \int_{\IR^d} \left\{
        \int_{0}^{N} \int_u^N g(x,v_1) (g(y,u)-g(x,u)) dv_1 du
    \right\} q^m_{\theta}(x,dy,\eta)
    , \eta(dx) \bigg \rangle .
\end{align*}
\normalsize
For the second generator, we have
\footnotesize
\begin{align*}
&
    \int A^N_2 f(\lp) \Gamma^N(\eta, d\lp)
\\ &\qquad =
    F_g^N(\eta) \bigg\langle \frac{c_\theta(x,\eta)}{\hat g(x)} \int_0^N g(x, u) \Bigg\{
        \int_u^N\Bigg(\int_{\IR^d}
        \left(\frac{g(y,v_1)+g(x,v_1)}{2}-1\right) q^m_{\theta}(x,dy,\eta)
        \Bigg)dv_1 \Bigg\}
    , \eta(dx) \bigg\rangle
\\ &\qquad =
    F_g^N(\eta) \bigg\langle \frac{c_\theta(x,\eta)}{\hat g(x)}
    \int_{\IR^d}\Bigg\{ \int_{0}^{N}
    \int_u^N g(x,u) \left(
        g(y,v_1)+g(x,v_1)-2
    \right) dv_1 du\Bigg\} q^m_{\theta}(x,dy,\eta)
    , \eta(dx) \bigg\rangle .
\end{align*}  
\normalsize
For the third generator we have that, since $g(x,N) = 1$,
\begin{align*}
&
    \int A^N_3f(\xi)\Gamma^N(\eta,d\lp)
\\ & \qquad =
    F_g^N(\eta) \bigg\langle \frac{1}{\hat g(x)}
    \int_0^N
        \left( c_\theta(x,\eta) u^2 - b_\theta(x,\eta) u \right) \partial_u g(x,u)
    du
    , \eta(dx) \bigg\rangle 
\\ & \qquad =
    F_g^N(\eta) \bigg\langle \frac{1}{\hat g(x)}
    \int_0^N
        \left( b_\theta(x,\eta) - 2 c_\theta(x,\eta) u \right) (g(x,u) - 1)
    du
    , \eta(dx) \bigg\rangle .
\end{align*}

Note that $2 \int_0^N \int_u^N g(x,v_1) g(y,u) dv_1 du = N^2 \hat g(x) \hat g(y)$,
and so
\begin{align*}
&
    \int_{0}^{N} \int_u^N g(x,v_1) (g(y,u)-g(x,u)) dv_1 du
    + \int_{0}^{N} \int_u^N g(x,u) \left(g(y,v_1)+g(x,v_1)-2\right)dv_1 du
\\ &\qquad = 
    N^2 \hat{g}(x)(\hat{g}(y) - 2) + 2\int_{0}^{N} u g(x,u) du.
\end{align*}

Combining the last equations,
and using the fact that
$N c_\theta(x,\eta) - b_\theta(x,\eta) = \theta \mu_\theta(x,\eta)$,
we have 
\begin{align*}
&
    \int \Bigg( A^N_1 f(\xi) + A^N_2 f(\xi) + A^N_3 f(\xi) \Bigg) \Gamma^N(\eta,d\lp)
\\ &\qquad =
    F_g^N(\eta) \bigg\langle
        c_\theta(x,\eta) N^2 \int_{\IR^d} (\hat g(y) - 2) q^m(x,dy,\eta)
        +
        \frac{1}{\hat g(x)} c_\theta(x,\eta) \int_0^N 2u du
\\ &\qquad \qquad \qquad {}
        +
        \frac{1}{\hat g(x)} b_\theta(x,\eta) \int_0^N (g(x,u) - 1) du
    , \eta(dx) \bigg\rangle 
\\ &\qquad =
    F_g^N(\eta) \bigg\langle
        c_\theta(x,\eta) N^2 \int_{\IR^d} (\hat g(y) - 1) q^m(x,dy,\eta)
        +
        N^2 c_\theta(x,\eta) \left(\frac{1}{\hat g(x)} - 1\right)
\\ &\qquad \qquad \qquad {}
        +
        N b_\theta(x,\eta) \left(1 - \frac{1}{\hat g(x)}\right)
    , \eta(dx) \bigg\rangle 
\\ &\qquad =
    F_g^N(\eta) N \bigg\langle
        N c_\theta(x,\eta) \int_{\IR^d} (\hat g(y) - 1) q^m(x,dy,\eta)
        +
        \theta \mu_\theta(x,\eta) \left(\frac{1}{\hat g(x)} - 1\right)
    , \eta(dx) \bigg\rangle  .
\end{align*}
This matches equation~\eqref{eqn:pgen_defn_again}, as desired,
because $N c_\theta(x,\eta) q_m(x,dy,\eta) = \theta \gamma(x,\eta) q_\theta(x,dy)$.
\end{proof}

>>>>

The following appears as Theorem A.2 in \citet{etheridge/kurtz:2019},
specialized slightly here to the case that the processes 
are c\`adl\`ag and have no fixed points of discontinuity.
For an $S_0$-valued, measurable process $Y$, $\hat{\mathcal{F}}^Y_t$
denotes the completion of the $\sigma$-algebra generated by
$Y(0)$ and $\{\int_0^r h(Y(s)) ds, r \le t, h \in B(S_0)\}$.
Also, let $D_S[0,\infty)$ denote the space of c\`adl\`ag, $S$-valued functions
with the Skorohod topology, and $M_S[0,\infty)$ the space of Borel measurable functions
from $[0,\infty)$ to $S$,
topologized by convergence in Lesbegue measure.
For other definitions see \citet{etheridge/kurtz:2019}.

@@

In this section we verify one of the conditions of the Markov Mapping Theorem,
namely, that ``integrating out levels'' in the generator of the lookdown process
we obtain the generator of the projected process.
In the notation of the theorem,
we are verifying that $C$ defined in~\eqref{eqn:C_defn} is in fact $\Pgen^N$
(if defined with $A^N$) or $\Pgen^\infty$ (if defined with $A$).
We will work with test functions of the form
\begin{equation} \label{eqn:f_defn}
    f(\lp) = \prod_{(x, u) \in \lp} g(x, u) = \exp\left(\langle \log g, \lp \rangle \right) ,
\end{equation}
where $0 \le g \le 1$ and $g(x,u) = 1$ for all $u \ge u_g$ for some $u_g < \infty$.
Furthermore, recall that $\kappa^N(\lp)(\cdot) = \lp(\cdot \times [0, N)) / N$
is the ``spatial projection operator'',
and define the transition function $\Gamma^N : \measures \to \mathcal{M}(\IR^d \times [0,N))$
so that for $\eta \in \measures$, if $\hat g_N(x) = \int_0^N g(x, u) du / N$, then
\begin{align*}
    F^N_g(\eta)
    &:=
    \int f(\lp) \Gamma^N(\eta, d\lp) \\
    &=
    \exp\left(
        N \left\langle
            \log \frac{1}{N} \int_0^N g(x, u) du, \eta(dx)
        \right\rangle
    \right) \\
    &=
    \exp\left(
        N \left\langle \log \hat g_N(x), \eta(dx) \right\rangle
    \right) ,
\end{align*}
i.e., $\Gamma^N$ assigns independent labels on $[0, N]$ to each of the points in $\eta$.
It follows from Lemma~\ref{def: MP definition of limit} that for test functions of this form
the generator of $\eta^N_t$ is
\begin{align} \label{eqn:pgen_defn_again}
    \begin{split}
    \Pgen^N F^N_g(\eta)
    =
    F^N_g(\eta)
        N \theta & \bigg\langle
        \gamma(x, \eta) \int r(z, \eta) \left( \hat g_N(z) - 1 \right) q_\theta(x, dz)
    \\ &\qquad \qquad {}
        +
        \mu_\theta(x, \eta) \left( \frac{1}{\hat g_N(x)} - 1 \right)
        ,
        \eta(dx)
    \bigg\rangle .
    \end{split}
\end{align}
(Note that $f$ here differs from the $f$ used in Lemma~\ref{def: MP definition of limit}
so as to agree with standard usage in the literature on lookdown processes.)
The generator of $\lp^N_t$ is $A^N$, defined in equation~\eqref{eqn:lookdown_generator}.

@@

We now integrate each piece against $\Gamma^N$.
First note that by the product form of $f$,
$$
    \int f(\lp) \sum_{(x,u)\in\lp} \frac{\ell(x,u)}{g(x,u)} \Gamma^N(\eta,d\lp)
    =
    F_g^N(\eta) \bigg\langle
        \frac{1}{N \hat g_N(x)} \int_0^N \ell(x,u) du,
        N \eta(dx)
    \bigg\rangle .
$$
Therefore,
\footnotesize
\begin{align*}
&
    \int A^N_1f(\lp) \Gamma^N(\eta,d\lp) 
\\ &\qquad =
    F_g^N(\eta) \bigg\langle \frac{c_\theta(x,\eta)}{\hat g_N(x)} \int_0^N \Bigg\{
        \int_u^N \Bigg(
            \frac{1}{2} g(x,v_1) \int_{\IR^d}(g(y,u)-g(x,u))q^m_{\theta}(x,dy,\eta)
        \Bigg) dv_1
    \Bigg\} du, \eta(dx) \bigg \rangle
\\ &\qquad =
    F_g^N(\eta) \bigg\langle \frac{c_\theta(x,\eta)}{\hat g_N(x)} \Bigg\{
    \int_{\IR^d} \left\{
        \int_{0}^{N} \int_u^N g(x,v_1) (g(y,u)-g(x,u)) dv_1 du
    \right\} q^m_{\theta}(x,dy,\eta)
    , \eta(dx) \bigg \rangle .
\end{align*}
\normalsize
For the second generator, we have
\footnotesize
\begin{align*}
&
    \int A^N_2 f(\lp) \Gamma^N(\eta, d\lp)
\\ &\qquad =
    F_g^N(\eta) \bigg\langle \frac{c_\theta(x,\eta)}{\hat g_N(x)} \int_0^N g(x, u) \Bigg\{
        \int_u^N\Bigg(\int_{\IR^d}
        \left(\frac{g(y,v_1)+g(x,v_1)}{2}-1\right) q^m_{\theta}(x,dy,\eta)
        \Bigg)dv_1 \Bigg\}
    , \eta(dx) \bigg\rangle
\\ &\qquad =
    F_g^N(\eta) \bigg\langle \frac{c_\theta(x,\eta)}{\hat g_N(x)}
    \int_{\IR^d}\Bigg\{ \int_{0}^{N}
    \int_u^N g(x,u) \left(
        g(y,v_1)+g(x,v_1)-2
    \right) dv_1 du\Bigg\} q^m_{\theta}(x,dy,\eta)
    , \eta(dx) \bigg\rangle .
\end{align*}  
\normalsize
For the third generator we have that
\begin{align*}
&
    \int A^N_3f(\xi)\Gamma^N(\eta,d\lp)
\\ & \qquad =
    F_g^N(\eta) \bigg\langle \frac{1}{\hat g_N(x)}
    \int_0^N
        \left( c_\theta(x,\eta) u^2 - b_\theta(x,\eta) u \right) \partial_u g(x,u)
    du
    , \eta(dx) \bigg\rangle 
\\ & \qquad =
    F_g^N(\eta) \bigg\langle \frac{1}{\hat g_N(x)}
    \int_0^N
        \left( b_\theta(x,\eta) - 2 c_\theta(x,\eta) u \right) (g(x,u) - 1)
    du
    , \eta(dx) \bigg\rangle .
\end{align*}

Note that $2 \int_0^N \int_u^N g(x,v_1) g(y,u) dv_1 du = N^2 \hat g_N(x) \hat g_N(y)$,
and so
\begin{align*}
&
    \int_{0}^{N} \int_u^N g(x,v_1) (g(y,u)-g(x,u)) dv_1 du
    + \int_{0}^{N} \int_u^N g(x,u) \left(g(y,v_1)+g(x,v_1)-2\right)dv_1 du
\\ &\qquad = 
    N^2 \hat{g_N}(x)(\hat{g_N}(y) - 2) + 2\int_{0}^{N} u g(x,u) du.
\end{align*}

Combining the last equations,
and using the fact that
$N c_\theta(x,\eta) - b_\theta(x,\eta) = \theta \mu_\theta(x,\eta)$,
we have 
\begin{align*}
&
    \int \Bigg( A^N_1 f(\xi) + A^N_2 f(\xi) + A^N_3 f(\xi) \Bigg) \Gamma^N(\eta,d\lp)
\\ &\qquad =
    F_g^N(\eta) \bigg\langle
        c_\theta(x,\eta) N^2 \int_{\IR^d} (\hat g_N(y) - 2) q^m(x,dy,\eta)
        +
        \frac{1}{\hat g_N(x)} c_\theta(x,\eta) \int_0^N 2u du
\\ &\qquad \qquad \qquad {}
        +
        \frac{1}{\hat g_N(x)} b_\theta(x,\eta) \int_0^N (g(x,u) - 1) du
    , \eta(dx) \bigg\rangle 
\\ &\qquad =
    F_g^N(\eta) \bigg\langle
        c_\theta(x,\eta) N^2 \int_{\IR^d} (\hat g_N(y) - 1) q^m(x,dy,\eta)
        +
        N^2 c_\theta(x,\eta) \left(\frac{1}{\hat g_N(x)} - 1\right)
\\ &\qquad \qquad \qquad {}
        +
        N b_\theta(x,\eta) \left(1 - \frac{1}{\hat g_N(x)}\right)
    , \eta(dx) \bigg\rangle 
\\ &\qquad =
    F_g^N(\eta) N \bigg\langle
        N c_\theta(x,\eta) \int_{\IR^d} (\hat g_N(y) - 1) q^m(x,dy,\eta)
        +
        \theta \mu_\theta(x,\eta) \left(\frac{1}{\hat g_N(x)} - 1\right)
    , \eta(dx) \bigg\rangle  .
\end{align*}
This matches equation~\eqref{eqn:pgen_defn_again}, as desired,
because $N c_\theta(x,\eta) q_m(x,dy,\eta) = \theta \gamma(x,\eta) q_\theta(x,dy)$.
\end{proof}
